---
title: "1DP_00_visium"
author: "Dominika Dostalova + Zuzana Matusova"
date: "2025-08-13"
output: html_document
editor_options: 
  chunk_output_type: console
---

### Overview
This script generates the data summaries required to visualize the proportion of *Gfap*-High reactive astrocytic spots via barplots, corresponding to the downstream analysis in `2V_00_visium.Rmd`.

### Data Provenance & Reproducibility
1. **Source Study:** This analysis utilizes a publicly available spatial transcriptomics dataset originally published by Zucha et al. (2024) (https://doi.org/10.1073/pnas.2404203121).
2. **Anatomical Focus:** Restricting analysis to the **Cortex** and **Stroke Lesion core** to directly match our single-cell RNA-seq datasets.
3. **Identification of *Gfap*-High Astrocytic Spots:** Spots positive for *Aldh1l1* expression ($Aldh1l1 > 0$) are dynamically binned into a "High-Expressing" reactive class if their *Gfap* expression exceeds the mean of the baseline control spots.


Create folders for saving objects
```{r Folders, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Define folders
folders <- c("ws", "ws/05_visium", "results", "results/05_visium")

# Create folders if they don't exist
lapply(folders, function(f) if (!dir.exists(f)) dir.create(f, recursive = TRUE))
```

```{r Load libraries and scripts, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
suppressMessages(library("Matrix"))
suppressMessages(library("DropletUtils"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("SeuratObject"))
suppressMessages(library("clustree"))
suppressMessages(library("DoubletFinder"))
suppressMessages(library("scProportionTest"))
suppressMessages(library("gridExtra"))
suppressMessages(library("dittoSeq"))
suppressMessages(library("dplyr"))
suppressMessages(library("SoupX"))
suppressMessages(library("ComplexHeatmap"))
suppressMessages(library("ggridges"))
suppressMessages(library("SeuratWrappers"))
suppressMessages(library("openxlsx"))
suppressMessages(library("EnhancedVolcano"))
suppressMessages(library("ggrepel"))
suppressMessages(library("clusterProfiler"))
suppressMessages(library("enrichplot"))
suppressMessages(library("AnnotationDbi"))
suppressMessages(library("org.Mm.eg.db"))

plots <- list()
col_list <- list()
orders <- list()
source("code/0F_01_colours_orders.R", local = F, echo = F, print.eval = F)
source("code/0F_02_general_functions.R", local = F, echo = F, print.eval = F)
gene_symbol <- read.table("data/all_genes.csv", header = TRUE, sep = ",")
```

Load Visium data
```{r seurat_object, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}}
seurat_object <- readRDS("data/spatial_seurat_1DP_13_nygen.rds")
```

Identification of Alhdhl1l and Gfap+ spots 
```{r visium_Gfap, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}

# Visium Spatial Gfap+ cells
seurat_object <- subset(seurat_object, subset = BrainAreas %in% c("Cortex","Lesion"))

# Extract metadata
meta <- seurat_object@meta.data

# Define gene
gene_name <- c("Gfap","Aldh1l1")

# Extract normalized expression
expr_data <- FetchData(seurat_object, vars = gene_name, layer = "data")  # This includes all cells
expr_data$sample <- seurat_object$Condition # Ensure metadata column matches
expr_data <- expr_data %>% filter(.[[gene_name[2]]] > 0) # all Aldh1l1+ cells

# Classify cells as Gfap-positive if expression > 0
expr_data$Gene_Positive <- expr_data[[gene_name[1]]] > 0 #Gfap+ cells from all Aldh1l1+ cells

#Gfap+ cells from all Aldh1l1+ cells with regulation for higher expression of Gfap+ cells than in controls
mean_expression <- expr_data %>%
  group_by(sample) %>%
  summarise(mean_expr = mean(.data[[gene_name[1]]][.data[[gene_name[1]]] > 0], na.rm = TRUE))

# Classify cells as "high-expressing" (above control threshold)
expr_data$Gene_Positive <- expr_data[[gene_name[1]]] > as.numeric(mean_expression[mean_expression$sample == "Ctrl","mean_expr"])


cell_summary <- expr_data %>%
  group_by(sample) %>%
  summarise(
    Total_Cells = n(),  # Total number of cells
    High_Expressing_Cells = sum(Gene_Positive)
  )

cell_summary <- cell_summary %>%
  mutate(Percent_Gene_Positive = round((High_Expressing_Cells / Total_Cells) * 100, 1))
cell_summary$sample <- factor(cell_summary$sample, levels = c("Ctrl", "1DPI", "3DPI", "7DPI"))

View(cell_summary)

saveRDS(cell_summary, "ws/05_visium/1DP_05_Gfap_summary_visium.rds")
```
