---
title: "1DP_03_astro_DEA_GO"
author: "Dominika Dostalova + Zuzana Matusova"
date: "2026-05-13"
output: html_document
editor_options: 
  chunk_output_type: console
---
# DEA analysis 
This notebook handles cell-count balancing, pairwise differential expression analysis, and downstream Gene Ontology (GO) biological process enrichment to evaluate transcriptomic variations between the WorthGentle and Miltenyi isolation protocols in MCAO-derived reactive astrocytes.

```{r packages}
suppressMessages(library("DropletUtils"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("SeuratObject"))
suppressMessages(library("clustree"))
suppressMessages(library("DoubletFinder"))
suppressMessages(library("scProportionTest"))
suppressMessages(library("gridExtra"))
suppressMessages(library("dittoSeq"))
suppressMessages(library("dplyr"))
suppressMessages(library("SoupX"))
suppressMessages(library("ComplexHeatmap"))
suppressMessages(library("ggridges"))
suppressMessages(library("SeuratWrappers"))
suppressMessages(library("openxlsx"))
suppressMessages(library("EnhancedVolcano"))
suppressMessages(library("ggrepel"))
suppressMessages(library("clusterProfiler"))
suppressMessages(library("enrichplot"))
suppressMessages(library("AnnotationDbi"))
suppressMessages(library("org.Mm.eg.db"))
suppressMessages(library("glmGamPoi"))
suppressMessages(library("future"))
sessionInfo()

plots <- list()
col_list <- list()
orders <- list()
source("code/0F_01_colours_orders.R", local = F, echo = F, print.eval = F)
source("code/0F_02_general_functions.R", local = F, echo = F, print.eval = F)
source("code/0F_03_label_names.R", local = F, echo = F, print.eval = F)
gene_symbol <- read.table("data/all_genes.csv", header = TRUE, sep = ",")
```

Create folders for saving objects
```{r Folders, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Define folders
folders <- c("ws", "ws/03_astro_DEA_GO", "results", "results/03_astro_DEA_GO", "results/supplements")

# Create folders if they don't exist
lapply(folders, function(f) if (!dir.exists(f)) dir.create(f, recursive = TRUE))
```

```{r}
seurat_object <- readRDS("ws/02_astro_subset/1DP_02_seurat_annotated_astro.rds")
table(seurat_object$SampleName, seurat_object$seurat_annotated_astro)
```

#################################################
# DEA ANALYSIS HOM vs REAC in ISOLATION PROTOCOLS
#################################################

```{r}
################################################################################
# PART 1: RE-RUN PAIRWISE DEA AND SAVE ALL TESTED GENES (BACKGROUND)
################################################################################

# 1. Load the pre-annotated astrocyte Seurat object
seurat_object <- readRDS("ws/02_astro_subset/1DP_02_seurat_annotated_astro.rds")

# 2. Subset down to MCAO Reactive Astrocytes for Miltenyi and WorthGentle
mcao_reac_seu <- subset(
  seurat_object, 
  subset = Condition == "MCAO" & seurat_annotated_astro == "Reac" & Isolation_pub %in% c("Miltenyi", "WorthGentle")
)
mcao_reac_seu$Isolation_pub <- droplevels(mcao_reac_seu$Isolation_pub)

# 3. Downsample to perfectly balance cell counts
min_cells_reac <- min(table(mcao_reac_seu$Isolation_pub))
Idents(mcao_reac_seu) <- "Isolation_pub"
set.seed(142) 
mcao_reac_balanced <- subset(mcao_reac_seu, downsample = min_cells_reac)

# 4. Run FindMarkers WITHOUT a logfc or p-value cutoff to capture all genes
DefaultAssay(mcao_reac_balanced) <- "RNA"
reac_miltenyi_vs_gentle_all_genes <- FindMarkers(
  object          = mcao_reac_balanced,
  ident.1         = "WorthGentle",
  ident.2         = "Miltenyi",
  logfc.threshold = 0,      # CRITICAL: 0 catches non-regulated background genes
  min.pct         = 0.10,   # Lowered to 10% to catch lower-expressed background genes
  test.use        = "wilcox"
)

# 5. Format and annotate the complete background matrix
reac_miltenyi_vs_gentle_all_genes$gene <- rownames(reac_miltenyi_vs_gentle_all_genes)
reac_miltenyi_vs_gentle_all_genes$gene_symbol <- ENSEMBL_to_Symbol(reac_miltenyi_vs_gentle_all_genes$gene)
reac_miltenyi_vs_gentle_all_genes$comparison  <- "WorthGentle_Reac_vs_Miltenyi_Reac_in_MCAO"

# 6. Save the UNFILTERED background dataset
saveRDS(reac_miltenyi_vs_gentle_all_genes, "ws/03_astro_DEA_GO/1DP_03_DEA_MCAO_Reac_WorthGentle_vs_Miltenyi.rds")
message("Saved complete background matrix. Total genes: ", nrow(reac_miltenyi_vs_gentle_all_genes))

#Excel

df_up   <- reac_miltenyi_vs_gentle_all_genes %>% filter(avg_log2FC > 0) %>% arrange(p_val_adj)
df_down <- reac_miltenyi_vs_gentle_all_genes %>% filter(avg_log2FC < 0) %>% arrange(p_val_adj)

#Initialize Excel Workbook
wb <- createWorkbook()

# --- MINIMAL BASIC STYLES ---
basic_header <- createStyle(textDecoration = "bold", halign = "center") 
p_val_style  <- createStyle(numFmt = "0.00E+00")
pct_style    <- createStyle(numFmt = "0.0%")
fc_style     <- createStyle(numFmt = "0.00")

# Minimal function to write tabs
populate_dea_sheet <- function(wb, sheet_name, df_data) {
  addWorksheet(wb, sheetName = sheet_name, gridLines = TRUE)
  
  # Select and order core Seurat metrics
  export_df <- df_data %>% 
    dplyr::select(gene, gene_symbol, avg_log2FC, pct.1, pct.2, p_val, p_val_adj, comparison)
  
  # Write data to sheet (Row 1 will be the headers)
  writeData(wb, sheet = sheet_name, x = export_df, withFilter = TRUE)
  
  # Apply basic formatting rules
  addStyle(wb, sheet = sheet_name, style = basic_header, rows = 1, cols = 1:8)
  addStyle(wb, sheet = sheet_name, style = fc_style,     rows = 2:(nrow(export_df)+1), cols = 3)
  
  # --- FIXED: Added gridExpand = TRUE for multi-row + multi-column styling ---
  addStyle(wb, sheet = sheet_name, style = pct_style,    rows = 2:(nrow(export_df)+1), cols = 4:5, gridExpand = TRUE)
  addStyle(wb, sheet = sheet_name, style = p_val_style,  rows = 2:(nrow(export_df)+1), cols = 6:7, gridExpand = TRUE)
  
  # Auto-fit columns to avoid truncated names or text wrapping issues
  setColWidths(wb, sheet = sheet_name, cols = 1:8, widths = "auto")
}

#Generate both directional data sheets
populate_dea_sheet(wb, "WorthGentle_UP", df_up)
populate_dea_sheet(wb, "Miltenyi_UP", df_down)

#Save final styled workbook
saveWorkbook(wb, "results/supplements/1DP_03_DEG_MCAO_Reac_Astro_MCAO_WorthGentle_vs_Miltenyi.xlsx", overwrite = TRUE)
```

####################################################
# GO ANALYSIS  REAC MILTENYI Miltenyi vs WorthGenle
####################################################

```{r}
# ---  Load the New Pairwise DEA Data ---
dea_data <- readRDS("ws/03_astro_DEA_GO/1DP_03_DEA_MCAO_Reac_WorthGentle_vs_Miltenyi.rds")

# Strip Ensembl trailing tags safely
dea_data <- dea_data %>%
  mutate(gene_clean = gsub("\\..*", "", gene))

# ---  Separate Genes by Direction of Fold Change ---
# Positive fold change = Enriched in WorthGentle (ident.1)
genes_up_worth <- dea_data %>%
  filter(p_val_adj < 0.05 & avg_log2FC > 0.65) %>%
  pull(gene_clean)

# Negative fold change = Enriched in Miltenyi (ident.2)
genes_up_miltenyi <- dea_data %>%
  filter(p_val_adj < 0.05 & avg_log2FC < -0.65) %>%
  pull(gene_clean)

message("Genes heading to WorthGentle High GO: ", length(genes_up_worth))
message("Genes heading to Miltenyi High GO: ", length(genes_up_miltenyi))


################################################################################
# STEP 4: RUN GENE ONTOLOGY ENRICHMENT (BIOLOGICAL PROCESSES)
################################################################################

# --- A. GO for WorthGentle Enriched Pathways ---
go_worth <- enrichGO(
  gene          = genes_up_worth,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)

if (!is.null(go_worth) && nrow(as.data.frame(go_worth)) > 0) {
  go_worth_simple <- clusterProfiler::simplify(go_worth)
  saveRDS(go_worth_simple, "ws/03_astro_DEA_GO/1DP_03_GO_WorthGentle_UP.rds")
  message("Successfully saved WorthGentle UP GO terms.")
} else {
  message("WARNING: No significant GO terms found for WorthGentle.")
}

# --- B. GO for Miltenyi Enriched Pathways ---
go_miltenyi <- enrichGO(
  gene          = genes_up_miltenyi,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)

if (!is.null(go_miltenyi) && nrow(as.data.frame(go_miltenyi)) > 0) {
  go_miltenyi_simple <- clusterProfiler::simplify(go_miltenyi)
  saveRDS(go_miltenyi_simple, "ws/03_astro_DEA_GO/1DP_03_GO_Miltenyi_UP.rds")
  message("Successfully saved Miltenyi UP GO terms.")
} else {
  message("WARNING: No significant GO terms found for Miltenyi.")
}

#Excel
# ==========================================
# STEP 1: LOAD AND CONVERT GO OBJECTS TO READABLE SYMBOLS
# ==========================================
go_worth_simple   <- readRDS("ws/03_astro_DEA_GO/1DP_03_GO_WorthGentle_UP.rds")
go_miltenyi_simple <- readRDS("ws/03_astro_DEA_GO/1DP_03_GO_Miltenyi_UP.rds")

go_worth_readable    <- setReadable(go_worth_simple, OrgDb = org.Mm.eg.db, keyType = "ENSEMBL")
go_miltenyi_readable <- setReadable(go_miltenyi_simple, OrgDb = org.Mm.eg.db, keyType = "ENSEMBL")

go_worth_df    <- as.data.frame(go_worth_readable)
go_miltenyi_df <- as.data.frame(go_miltenyi_readable)

if(nrow(go_worth_df) > 0)    go_worth_df    <- go_worth_df %>% rename(geneSymbol = geneID)
if(nrow(go_miltenyi_df) > 0) go_miltenyi_df <- go_miltenyi_df %>% rename(geneSymbol = geneID)

# ==========================================
# STEP 2: INITIALIZE THE EXCEL WORKBOOK
# ==========================================
wb <- createWorkbook()

# ==========================================
# STEP 3: FORMAT AND EXPORT DATA TO TABS
# ==========================================

# --- Tab 1: WorthGentle Enriched ---
if (nrow(go_worth_df) > 0) {
  sub_data_worth <- go_worth_df %>%
    dplyr::select("ID", "Description", "GeneRatio", "BgRatio", "pvalue", "p.adjust", "qvalue", "geneSymbol", "Count") %>%
    arrange(p.adjust)
  
  sheet_name_worth <- "WorthGentle_UP"
  addWorksheet(wb, sheetName = sheet_name_worth)
  writeData(wb, sheet = sheet_name_worth, x = sub_data_worth, withFilter = TRUE)
} else {
  message("No significant WorthGentle terms to write.")
}

# --- Tab 2: Miltenyi Enriched ---
if (nrow(go_miltenyi_df) > 0) {
  sub_data_miltenyi <- go_miltenyi_df %>%
    dplyr::select("ID", "Description", "GeneRatio", "BgRatio", "pvalue", "p.adjust", "qvalue", "geneSymbol", "Count") %>%
    arrange(p.adjust)
  
  sheet_name_miltenyi <- "Miltenyi_UP"
  addWorksheet(wb, sheetName = sheet_name_miltenyi)
  writeData(wb, sheet = sheet_name_miltenyi, x = sub_data_miltenyi, withFilter = TRUE)
} else {
  message("No significant Miltenyi terms to write.")
}

# ==========================================
# STEP 4: SAVE WORKBOOK
# ==========================================

saveWorkbook(wb, "results/supplements/1DP_03_GO_BP_Reac_Astro_MCAO_WorthGentle_vs_Miltenyi.xlsx", overwrite = TRUE)
```