---
title: "1DP_03_astro_DEA_GO"
author: "Dominika Dostalova + Zuzana Matusova"
date: "2026-05-13"
output: html_document
editor_options: 
  chunk_output_type: console
---
# DEA analysis 
This notebook handles cell-count balancing, pairwise differential expression analysis, and downstream Gene Ontology (GO) biological process enrichment to evaluate transcriptomic variations between the WorthGentle and Miltenyi isolation protocols in MCAO-derived reactive astrocytes.

```{r packages}
suppressMessages(library("DropletUtils"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("SeuratObject"))
suppressMessages(library("clustree"))
suppressMessages(library("DoubletFinder"))
suppressMessages(library("scProportionTest"))
suppressMessages(library("gridExtra"))
suppressMessages(library("dittoSeq"))
suppressMessages(library("dplyr"))
suppressMessages(library("SoupX"))
suppressMessages(library("ComplexHeatmap"))
suppressMessages(library("ggridges"))
suppressMessages(library("SeuratWrappers"))
suppressMessages(library("openxlsx"))
suppressMessages(library("EnhancedVolcano"))
suppressMessages(library("ggrepel"))
suppressMessages(library("clusterProfiler"))
suppressMessages(library("enrichplot"))
suppressMessages(library("AnnotationDbi"))
suppressMessages(library("org.Mm.eg.db"))
suppressMessages(library("glmGamPoi"))
suppressMessages(library("future"))
sessionInfo()

plots <- list()
col_list <- list()
orders <- list()
source("code/0F_01_colours_orders.R", local = F, echo = F, print.eval = F)
source("code/0F_02_general_functions.R", local = F, echo = F, print.eval = F)
source("code/0F_03_label_names.R", local = F, echo = F, print.eval = F)
gene_symbol <- read.table("data/all_genes.csv", header = TRUE, sep = ",")
```

Create folders for saving objects
```{r Folders, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Define folders
folders <- c("ws", "ws/03_astro_DEA_GO", "results", "results/03_astro_DEA_GO", "results/supplements")

# Create folders if they don't exist
lapply(folders, function(f) if (!dir.exists(f)) dir.create(f, recursive = TRUE))
```

```{r}
seurat_object <- readRDS("ws/02_astro_subset/1DP_02_seurat_annotated_astro.rds")
table(seurat_object$SampleName, seurat_object$seurat_annotated_astro)
```

#################################################
# DEA ANALYSIS HOM vs REAC in ISOLATION PROTOCOLS
#################################################

```{r}
################################################################################
# PART 1: RE-RUN PAIRWISE DEA AND SAVE ALL TESTED GENES (BACKGROUND)
################################################################################

# 1. Load the pre-annotated astrocyte Seurat object
seurat_object <- readRDS("ws/02_astro_subset/1DP_02_seurat_annotated_astro.rds")

# 2. Subset down to MCAO Reactive Astrocytes for Miltenyi and WorthGentle
mcao_reac_seu <- subset(
  seurat_object, 
  subset = Condition == "MCAO" & seurat_annotated_astro == "Reac" & Isolation_pub %in% c("Miltenyi", "WorthGentle")
)
mcao_reac_seu$Isolation_pub <- droplevels(mcao_reac_seu$Isolation_pub)

# 3. Downsample to perfectly balance cell counts
#min_cells_reac <- min(table(mcao_reac_seu$Isolation_pub))
Idents(mcao_reac_seu) <- "Isolation_pub"
#set.seed(142) 
#mcao_reac_balanced <- subset(mcao_reac_seu, downsample = min_cells_reac)
mcao_reac_balanced <- mcao_reac_seu

# 4. Run FindMarkers WITHOUT a logfc or p-value cutoff to capture all genes
DefaultAssay(mcao_reac_balanced) <- "RNA"
reac_miltenyi_vs_gentle_all_genes <- FindMarkers(
  object          = mcao_reac_balanced,
  ident.1         = "WorthGentle",
  ident.2         = "Miltenyi",
  logfc.threshold = 0,      # CRITICAL: 0 catches non-regulated background genes
  min.pct         = 0.10,   # Lowered to 10% to catch lower-expressed background genes
  test.use        = "wilcox"
)

# 5. Format and annotate the complete background matrix
reac_miltenyi_vs_gentle_all_genes$gene <- rownames(reac_miltenyi_vs_gentle_all_genes)
reac_miltenyi_vs_gentle_all_genes$gene_symbol <- ENSEMBL_to_Symbol(reac_miltenyi_vs_gentle_all_genes$gene)
reac_miltenyi_vs_gentle_all_genes$comparison  <- "WorthGentle_Reac_vs_Miltenyi_Reac_in_MCAO"

# 6. Save the UNFILTERED background dataset
saveRDS(reac_miltenyi_vs_gentle_all_genes, "ws/03_astro_DEA_GO/1DP_03_DEA_MCAO_Reac_WorthGentle_vs_Miltenyi.rds")
message("Saved complete background matrix. Total genes: ", nrow(reac_miltenyi_vs_gentle_all_genes))

#csv

df_up   <- reac_miltenyi_vs_gentle_all_genes %>% filter(avg_log2FC > 0) %>% arrange(p_val_adj)
df_down <- reac_miltenyi_vs_gentle_all_genes %>% filter(avg_log2FC < 0) %>% arrange(p_val_adj)

# ==========================================
# COMBINE AND PREPARE DEA DATA FOR CSV EXPORT
# ==========================================
# Combine both directions, ensuring the direction/comparison is clearly labeled
dea_combined <- bind_rows(
  df_up %>% mutate(direction = "WorthGentle_UP"),
  df_down %>% mutate(direction = "Miltenyi_UP")
) %>%
  # Select core metrics and order logically
  dplyr::select(
    gene, 
    gene_symbol, 
    avg_log2FC, 
    pct.1, 
    pct.2, 
    p_val, 
    p_val_adj, 
    comparison,
    direction
  ) %>%
  arrange(direction, p_val_adj, desc(abs(avg_log2FC)))

# ==========================================
# EXPORT AS CSV
# ==========================================
write.csv(
  x         = dea_combined,
  file      = "results/supplements/1DP_03_DEG_MCAO_Reac_Astro_MCAO_WorthGentle_vs_Miltenyi.csv",
  row.names = FALSE
)

message("Saved DEA CSV successfully. Total rows: ", nrow(dea_combined))
```

####################################################
# GO ANALYSIS  REAC MILTENYI Miltenyi vs WorthGenle
####################################################

```{r}
# ---  Load the New Pairwise DEA Data ---
dea_data <- readRDS("ws/03_astro_DEA_GO/1DP_03_DEA_MCAO_Reac_WorthGentle_vs_Miltenyi.rds")

# Strip Ensembl trailing tags safely
dea_data <- dea_data %>%
  mutate(gene_clean = gsub("\\..*", "", gene))

# ---  Separate Genes by Direction of Fold Change ---
# Positive fold change = Enriched in WorthGentle (ident.1)
genes_up_worth <- dea_data %>%
  filter(p_val_adj < 0.05 & avg_log2FC > 0.65) %>%
  pull(gene_clean)

# Negative fold change = Enriched in Miltenyi (ident.2)
genes_up_miltenyi <- dea_data %>%
  filter(p_val_adj < 0.05 & avg_log2FC < -0.65) %>%
  pull(gene_clean)

message("Genes heading to WorthGentle High GO: ", length(genes_up_worth))
message("Genes heading to Miltenyi High GO: ", length(genes_up_miltenyi))


################################################################################
# STEP 4: RUN GENE ONTOLOGY ENRICHMENT (BIOLOGICAL PROCESSES)
################################################################################

# --- A. GO for WorthGentle Enriched Pathways ---
go_worth <- enrichGO(
  gene          = genes_up_worth,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)

if (!is.null(go_worth) && nrow(as.data.frame(go_worth)) > 0) {
  go_worth_simple <- clusterProfiler::simplify(go_worth)
  saveRDS(go_worth_simple, "ws/03_astro_DEA_GO/1DP_03_GO_WorthGentle_UP.rds")
  message("Successfully saved WorthGentle UP GO terms.")
} else {
  message("WARNING: No significant GO terms found for WorthGentle.")
}

# --- B. GO for Miltenyi Enriched Pathways ---
go_miltenyi <- enrichGO(
  gene          = genes_up_miltenyi,
  OrgDb         = org.Mm.eg.db,
  keyType       = "ENSEMBL",
  ont           = "BP",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)

if (!is.null(go_miltenyi) && nrow(as.data.frame(go_miltenyi)) > 0) {
  go_miltenyi_simple <- clusterProfiler::simplify(go_miltenyi)
  saveRDS(go_miltenyi_simple, "ws/03_astro_DEA_GO/1DP_03_GO_Miltenyi_UP.rds")
  message("Successfully saved Miltenyi UP GO terms.")
} else {
  message("WARNING: No significant GO terms found for Miltenyi.")
}

#Excel
# ==========================================
# STEP 1: LOAD AND CONVERT GO OBJECTS TO READABLE SYMBOLS
# ==========================================
go_worth_simple   <- readRDS("ws/03_astro_DEA_GO/1DP_03_GO_WorthGentle_UP.rds")
go_miltenyi_simple <- readRDS("ws/03_astro_DEA_GO/1DP_03_GO_Miltenyi_UP.rds")

go_worth_readable    <- setReadable(go_worth_simple, OrgDb = org.Mm.eg.db, keyType = "ENSEMBL")
go_miltenyi_readable <- setReadable(go_miltenyi_simple, OrgDb = org.Mm.eg.db, keyType = "ENSEMBL")

go_worth_df    <- as.data.frame(go_worth_readable)
go_miltenyi_df <- as.data.frame(go_miltenyi_readable)

if(nrow(go_worth_df) > 0)    go_worth_df    <- go_worth_df %>% rename(geneSymbol = geneID)
if(nrow(go_miltenyi_df) > 0) go_miltenyi_df <- go_miltenyi_df %>% rename(geneSymbol = geneID)

# ==========================================
# STEP 2 & 3: COMBINE AND FORMAT GO DATA
# ==========================================
target_cols <- c("ID", "Description", "GeneRatio", "BgRatio", "pvalue", "p.adjust", "qvalue", "geneSymbol", "Count")

# Prepare WorthGentle terms
sub_data_worth <- data.frame()
if (nrow(go_worth_df) > 0) {
  sub_data_worth <- go_worth_df %>%
    dplyr::select(all_of(target_cols)) %>%
    mutate(Direction = "WorthGentle_UP")
}

# Prepare Miltenyi terms
sub_data_miltenyi <- data.frame()
if (nrow(go_miltenyi_df) > 0) {
  sub_data_miltenyi <- go_miltenyi_df %>%
    dplyr::select(all_of(target_cols)) %>%
    mutate(Direction = "Miltenyi_UP")
}

# Combine into a single structured data frame
go_combined <- bind_rows(sub_data_worth, sub_data_miltenyi) %>%
  dplyr::select(Direction, all_of(target_cols)) %>%
  arrange(Direction, p.adjust)

# ==========================================
# STEP 4: SAVE AS CSV
# ==========================================
if (nrow(go_combined) > 0) {
  write.csv(
    x         = go_combined,
    file      = "results/supplements/1DP_03_GO_BP_Reac_Astro_MCAO_WorthGentle_vs_Miltenyi.csv",
    row.names = FALSE
  )
  message("Saved GO CSV successfully. Total rows: ", nrow(go_combined))
} else {
  message("Warning: No significant GO terms found across either condition; CSV not written.")
}
```

```{r}
library(dplyr)
library(clusterProfiler)
library(org.Mm.eg.db)
library(AnnotationDbi)

# ==============================================================================
# STEP 1: LOAD DEA DATA & CLEAN ENSEMBL IDS
# ==============================================================================
dea_data <- readRDS("ws/03_astro_DEA_GO/1DP_03_DEA_MCAO_Reac_WorthGentle_vs_Miltenyi.rds")

# Strip Ensembl trailing version tags (e.g., ENSMUSG00000001234.5 -> ENSMUSG00000001234)
dea_data <- dea_data %>%
  mutate(gene_clean = gsub("\\..*", "", gene))

# ==============================================================================
# STEP 2: CATEGORIZE BY STATISTICAL SIGNIFICANCE & DIRECTION
# ==============================================================================
# 1. Non-significant / low fold change genes (filtered out before GO)
dea_not_significant <- dea_data %>%
  filter(p_val_adj >= 0.05 | abs(avg_log2FC) <= 0.65 | is.na(p_val_adj))

# 2. Significant candidate genes heading to GO
worth_candidates <- dea_data %>%
  filter(p_val_adj < 0.05 & avg_log2FC > 0.65)

miltenyi_candidates <- dea_data %>%
  filter(p_val_adj < 0.05 & avg_log2FC < -0.65)

cat("Total genes in background DEA:", nrow(dea_data), "\n")
cat("Excluded by DEA thresholds (p_adj >= 0.05 or |log2FC| <= 0.65):", nrow(dea_not_significant), "\n")
cat("WorthGentle significant candidate genes:", nrow(worth_candidates), "\n")
cat("Miltenyi significant candidate genes:", nrow(miltenyi_candidates), "\n\n")

# ==============================================================================
# STEP 3: FUNCTION TO IDENTIFY UNMAPPED / UNANNOTATED GENES IN ORG.MM.EG.DB
# ==============================================================================
audit_go_input_genes <- function(candidate_df, group_name) {
  ensembl_ids <- unique(candidate_df$gene_clean)
  
  # A. Check mapping to Entrez ID
  map_entrez <- AnnotationDbi::select(
    org.Mm.eg.db,
    keys     = ensembl_ids,
    keytype  = "ENSEMBL",
    columns  = c("ENTREZID", "SYMBOL")
  ) %>% distinct(ENSEMBL, .keep_all = TRUE)
  
  # Unmapped to Entrez
  unmapped_entrez <- map_entrez %>%
    filter(is.na(ENTREZID)) %>%
    pull(ENSEMBL)
  
  # B. Check mapping to GO Biological Process (BP) terms
  mapped_entrez_ids <- na.omit(map_entrez$ENTREZID)
  
  map_go <- AnnotationDbi::select(
    org.Mm.eg.db,
    keys     = mapped_entrez_ids,
    keytype  = "ENTREZID",
    columns  = c("GO", "ONTOLOGY")
  ) %>% filter(ONTOLOGY == "BP")
  
  entrez_with_go_bp <- unique(na.omit(map_go$ENTREZID))
  
  # Entrez IDs with NO Biological Process annotations
  entrez_no_go_bp <- setdiff(mapped_entrez_ids, entrez_with_go_bp)
  ensembl_no_go_bp <- map_entrez %>%
    filter(ENTREZID %in% entrez_no_go_bp) %>%
    pull(ENSEMBL)
  
  # Genes successfully entering the hypergeometric test
  ensembl_valid_go <- map_entrez %>%
    filter(ENTREZID %in% entrez_with_go_bp) %>%
    pull(ENSEMBL)
  
  # Construct status table for all candidates
  candidate_status <- candidate_df %>%
    mutate(
      GO_inclusion_status = case_when(
        gene_clean %in% unmapped_entrez   ~ "Excluded: No Entrez ID in org.Mm.eg.db",
        gene_clean %in% ensembl_no_go_bp  ~ "Excluded: No GO Biological Process annotation",
        gene_clean %in% ensembl_valid_go  ~ "Included in GO hypergeometric test",
        TRUE                              ~ "Excluded"
      )
    )
  
  cat(paste0("--- ", group_name, " Audit Summary ---\n"))
  cat("Total candidate DEGs:", nrow(candidate_df), "\n")
  cat("  -> Included in GO BP test:", length(ensembl_valid_go), "\n")
  cat("  -> Excluded (No Entrez ID):", length(unmapped_entrez), "\n")
  cat("  -> Excluded (No GO BP terms):", length(ensembl_no_go_bp), "\n\n")
  
  return(candidate_status)
}

# ==============================================================================
# STEP 4: EXECUTE AUDIT FOR BOTH PROTOCOLS
# ==============================================================================
worth_audit    <- audit_go_input_genes(worth_candidates, "WorthGentle UP")
miltenyi_audit <- audit_go_input_genes(miltenyi_candidates, "Miltenyi UP")

# ==============================================================================
# STEP 5: EXTRACT ONLY THE EXCLUDED GENES & SAVE TO CSV
# ==============================================================================
# Filter to genes that met DEA significance but could not be evaluated in GO
worth_excluded <- worth_audit %>%
  filter(GO_inclusion_status != "Included in GO hypergeometric test") %>%
  dplyr::select(gene, gene_clean, gene_symbol, avg_log2FC, p_val_adj, GO_inclusion_status)

miltenyi_excluded <- miltenyi_audit %>%
  filter(GO_inclusion_status != "Included in GO hypergeometric test") %>%
  dplyr::select(gene, gene_clean, gene_symbol, avg_log2FC, p_val_adj, GO_inclusion_status)

# Combine into a single export table
all_deg_go_excluded <- bind_rows(
  worth_excluded %>% mutate(Direction = "WorthGentle_UP"),
  miltenyi_excluded %>% mutate(Direction = "Miltenyi_UP")
)

# Export excluded candidate genes
write.csv(
  all_deg_go_excluded,
  file = "results/supplements/DEGs_excluded_from_GO_due_to_unmapped_annotations.csv",
  row.names = FALSE
)

message("Audit complete. Saved list of excluded candidate genes to: ")
message("results/supplements/DEGs_excluded_from_GO_due_to_unmapped_annotations.csv")
```

