---
title: "1DP_02_astro_subset"
author: "Dominika Dostalova + Zuzana Matusova"
date: "2025-07-24"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Seurat object only with astrocytes
This notebook contains the primary preprocessing, quality control, clustering, and population annotation pipeline for a subset of astrocytes in single-cell transcriptomic study evaluating different brain tissue isolation protocols.

Create folders for saving objects
```{r Folders, eval = TRUE, echo = FALSE, message = FALSE, warning=FALSE}
# Define folders
folders <- c("ws", "ws/02_astro_subset", "results", "results/02_astro_subset", "results/supplements")

# Create folders if they don't exist
lapply(folders, function(f) if (!dir.exists(f)) dir.create(f, recursive = TRUE))
```

Load packages and print sessionInfo to see package versions.
```{r packages}
suppressMessages(library("DropletUtils"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("SeuratObject"))
suppressMessages(library("clustree"))
suppressMessages(library("DoubletFinder"))
suppressMessages(library("scProportionTest"))
suppressMessages(library("gridExtra"))
suppressMessages(library("dittoSeq"))
suppressMessages(library("dplyr"))
suppressMessages(library("SoupX"))
suppressMessages(library("ComplexHeatmap"))
suppressMessages(library("ggridges"))
suppressMessages(library("SeuratWrappers"))
suppressMessages(library("openxlsx"))
suppressMessages(library("EnhancedVolcano"))
suppressMessages(library("ggrepel"))
suppressMessages(library("clusterProfiler"))
suppressMessages(library("enrichplot"))
suppressMessages(library("AnnotationDbi"))
suppressMessages(library("org.Mm.eg.db"))
suppressMessages(library("glmGamPoi"))
suppressMessages(library("tibble"))
sessionInfo()

plots <- list()
col_list <- list()
orders <- list()
source("code/0F_01_colours_orders.R", local = F, echo = F, print.eval = F)
source("code/0F_02_general_functions.R", local = F, echo = F, print.eval = F)
source("code/0F_03_label_names.R", local = F, echo = F, print.eval = F)
gene_symbol <- read.table("data/all_genes.csv", header = TRUE, sep = ",")

```


Astro only
```{r}
seurat_object <- readRDS("ws/01_all_cells/1DP_01_seurat_annotated_all_cells.rds")
```

```{r}
#Excluded Aldh1l1 neg cells from astrocytes

# 1. Ensure active ident is set to annotated cell types and subset Astrocytes
Idents(seurat_object) <- "seurat_annotated"
astro_cluster <- subset(seurat_object, idents = "Ast")

# 2. Set default assay to RNA
DefaultAssay(astro_cluster) <- "RNA"

# 3. Retrieve Ensembl ID for Aldh1l1
aldh1l1_id <- Symbol_to_ENSEMBL("Aldh1l1")

# 4. Fetch Aldh1l1 expression and metadata
expr_data_cond <- FetchData(
  astro_cluster, 
  vars = c(aldh1l1_id, "Isolation_pub", "Condition"), 
  layer = "data"
)

# Safely rename the gene expression column by matching its Ensembl ID name
colnames(expr_data_cond)[colnames(expr_data_cond) == aldh1l1_id] <- "Aldh1l1"

# 5. Summarise per protocol and condition
exclusion_by_cond <- expr_data_cond %>%
  group_by(Isolation_pub, Condition) %>%
  summarise(
    Total = n(),
    Retained = sum(Aldh1l1 > 0),
    Excluded = sum(Aldh1l1 == 0),
    Pct_Excluded = round((Excluded / Total) * 100, 2),
    .groups = "drop"
  )

cat("\n=== EXCLUSION BY PROTOCOL & CONDITION ===\n")
print(as.data.frame(exclusion_by_cond))

# 6. Summarise per isolation protocol overall
exclusion_by_protocol <- expr_data_cond %>%
  group_by(Isolation_pub) %>%
  summarise(
    Total_in_Ast = n(),
    Retained_Aldh1l1_pos = sum(Aldh1l1 > 0),
    Excluded_Aldh1l1_zero = sum(Aldh1l1 == 0),
    Pct_Excluded = round((Excluded_Aldh1l1_zero / Total_in_Ast) * 100, 2),
    .groups = "drop"
  )

cat("\n=== EXCLUSION BY PROTOCOL OVERALL ===\n")
print(as.data.frame(exclusion_by_protocol))
```

```{r subset}
# Subset
Idents(seurat_object) <- "seurat_annotated"
seurat_object <- subset(seurat_object, ident = c("Ast"))
expr_data <- FetchData(seurat_object, vars = Symbol_to_ENSEMBL("Aldh1l1"), layer = "data")
cells_select <- expr_data %>% filter(.[[1]] > 0) %>% row.names
seurat_object <- subset(seurat_object, cells = cells_select)
```


# normalize, scale, SCT, cluster
Normalization and scaling of the RNA assay.
```{r normalize}

DefaultAssay(seurat_object) <- "RNA"
all.genes.RNA <- rownames(seurat_object)
seurat_object <- NormalizeData(seurat_object, normalization.method = "LogNormalize", scale.factor = 10000)
seurat_object <- ScaleData(seurat_object, features = all.genes.RNA)

options(future.globals.maxSize = 8 * 1024^3)
set.seed(1234)
seurat_object <- SCTransform(seurat_object, verbose = TRUE, return.only.var.genes = TRUE, 
                             assay = "RNA", n_genes = NULL, variable.features.n = NULL, method = "glmGamPoi")
```


First, decide which assay will be used for PCA, UMAP, and clustering - integrated or SCT. In PCA, number of genes used can be reduced from all to e. g. only variable. Based on Elbow Plot decide, how many dimensions will be used in further analysis - how many is reasonable for explaining the most of the variability. Run UMAP on selected number of dimensions.
```{r PCA}
DefaultAssay(seurat_object) <- "SCT" # can be set to SCT or integrated


all.genes <- row.names(seurat_object)

set.seed(1234)
seurat_object <- RunPCA(seurat_object, npcs = 50, verbose = FALSE, features = all.genes)
print(ElbowPlot(seurat_object, ndims = 50, reduction = "pca"))

ndim <- 20

set.seed(1234)
seurat_object <- RunUMAP(seurat_object, reduction = "pca", dims = 1:ndim, verbose = TRUE)

set.seed(1234)
seurat_object <- FindNeighbors(seurat_object, reduction = "pca", dims = 1:ndim, verbose = TRUE)

rm(all.genes)
```


Proceed with clustering. Choose optimal resolution, here set to 0.5. {clustree} package can help decide, as it plots multiple resolutions and connections between the identified clusters. Based on this plot the most reasonable resolution can be selected (the one where relationships between clusters are clear and not intermixed).

```{r clustree}
# Clustree, to help with selection of optimal clustering resolution.

seurat_clustree <- seurat_object
seurat_clustree@meta.data <- seurat_clustree@meta.data %>% 
  dplyr::select(-contains("_snn_res."))
assay <- seurat_object@active.assay

# --- Define your exact custom resolutions here ---
custom_resolutions <- c(0.05, 0.075, 0.1, 0.15)

for (resol in custom_resolutions) {
  message("Calculating clusters for resolution: ", resol)
  set.seed(1234)
  seurat_clustree <- FindClusters(
    seurat_clustree, 
    resolution = resol, 
    graph.name = paste0(assay, "_snn"), 
    verbose = FALSE
  )
}

# Print the tree graph using your custom steps
print(clustree(seurat_clustree, prefix = paste0(assay, "_snn_res.")))
```

```{r clustering}
# set the optimal resolution based on clustree, prior knowledge, or randomly

res <- 0.075

set.seed(1234)
seurat_object <- FindClusters(seurat_object, resolution = res, verbose = TRUE)

# check DimPlot using the optimal resolution
DimPlot(seurat_object, label = T, label.box = T, repel = T) + NoLegend()
```

```{r}
# ==========================================
# COMPUTE CLUSTER MARKERS
# ==========================================
DefaultAssay(seurat_object) <- "RNA"
Idents(seurat_object) <- "seurat_clusters"
markers <- FindAllMarkers(seurat_object, min.pct = 0.8, logfc.threshold = 0.65, only.pos = TRUE, verbose = TRUE)

# Translate Ensembl gene IDs to symbols
markers$gene_symbol <- ENSEMBL_to_Symbol(markers$gene)
```


```{r}
# ==========================================
# ANNOTATE SEURAT METADATA
# ==========================================
# FIXED: Using uppercase "HOM" and "REAC" to guarantee flawless alignment with global project tracks
seurat_object@meta.data <- seurat_object@meta.data %>% 
  mutate(seurat_annotated_astro = factor(case_when(
    seurat_clusters == 0 ~ "Hom",
    seurat_clusters == 1 ~ "Reac",
    TRUE                 ~ as.character(seurat_clusters)
  )))

print("Metadata verification count:")
print(table(seurat_object$seurat_annotated_astro))

saveRDS(seurat_object, "ws/02_astro_subset/1DP_02_seurat_annotated_astro.rds")
```

#Supplement with cluster markers

```{r}
seurat_object <- readRDS("ws/02_astro_subset/1DP_02_seurat_annotated_astro.rds")
# ==========================================
# COMPUTE CLUSTER MARKERS
# ==========================================
DefaultAssay(seurat_object) <- "RNA"
Idents(seurat_object) <- "seurat_clusters"
markers <- FindAllMarkers(seurat_object, min.pct = 0.8, logfc.threshold = 0.65, only.pos = TRUE, verbose = TRUE)

# Translate Ensembl gene IDs to symbols
markers$gene_symbol <- ENSEMBL_to_Symbol(markers$gene)
```

```{r}
# ==========================================
# PREPARE DATA FRAMES FOR EXPORT
# ==========================================
# Safe, type-independent numeric matching
markers <- markers %>%
  mutate(cell_type_annotated = case_when(
    cluster == 0 ~ "Hom",
    cluster == 1 ~ "Reac",
    TRUE         ~ as.character(cluster)
  ))

# Define sheet / cluster factor order
if (exists("custom_sheet")) {
  valid_order <- custom_sheet[custom_sheet %in% c("Hom", "Reac")]
} else {
  valid_order <- c("Hom", "Reac")
}

# Filter, order, and select relevant columns
markers_clean <- markers %>%
  filter(cell_type_annotated %in% c("Hom", "Reac")) %>%
  mutate(cell_type_annotated = factor(cell_type_annotated, levels = valid_order)) %>%
  arrange(cell_type_annotated, p_val_adj, desc(avg_log2FC)) %>%
  dplyr::select(
    gene_symbol,
    gene,
    cell_type_annotated,
    p_val,
    avg_log2FC,
    pct.1,
    pct.2,
    p_val_adj
  )

# ==========================================
# SAVE AS CSV
# ==========================================
write.csv(
  x         = markers_clean,
  file      = "results/supplements/1DP_02_markers_astro_subset.csv",
  row.names = FALSE
)

message("Saved astrocyte subset markers CSV successfully. Total rows: ", nrow(markers_clean))
```


# Gfap+ cells - numbers
```{r Gfap_cells} 
gene_name <- Symbol_to_ENSEMBL("Gfap")
seurat_object <- readRDS("ws/02_astro_subset/1DP_02_seurat_annotated_astro.rds")

# Extract metadata
seurat_object$Isolation_pub <- factor(seurat_object$Isolation_pub, levels = c("Miltenyi", "WorthGentle","WorthMech"))

# Define gene
gene_name <- Symbol_to_ENSEMBL(c("Gfap"))

# Extract normalized expression
expr_data <- FetchData(seurat_object, vars = c(gene_name, "Isolation_pub", "Condition"), layer = "data")  # This includes all cells
expr_data_mcao <- expr_data %>% filter(Condition == "MCAO")

# Calculate Mean Expression for Gfap-positive MCAO cells
mean_expression <- expr_data_mcao %>%
  group_by(Isolation_pub) %>%
  summarise(mean_expr = mean(.data[[gene_name[1]]][.data[[gene_name[1]]] > 0], na.rm = TRUE))

# Calculate total cell counts and positive expressors within MCAO
cell_summary <- expr_data_mcao %>%
  group_by(Isolation_pub) %>%
  summarise(
    Total_Cells = n(),  # Total number of MCAO cells
    Expressing_Cells = sum(.data[[gene_name[1]]] > 0)
  )

# Compute percentage of Gfap+ cells out of all MCAO cells
cell_summary <- cell_summary %>%
  mutate(Percent_Gene_Positive = round((Expressing_Cells / Total_Cells) * 100, 1))


saveRDS(cell_summary, "ws/02_astro_subset/1DP_02_Gfap_cell_summary.rds")
```