---
title: "1DP_00_raw_data_to_seurat"
author: "Pavel Abaffy + Dominika Dostalova"
date: "2026-06-05"
output: html_document
---
#Introduction
This script outlines the pipeline used to process raw sequencing data into a Seurat object. Because the workflow spans multiple environments and programming languages, this script serves as the backbone documenting where and how the data are processed. It is divided into three major sections: pre-filtering the original FASTQ files, processing the obtained FASTQ files via the split-pipe pipeline, and generating the Seurat object.

########################################
# PRE-FILTERING THE ORIGINAL FASTQ FILES
########################################
The original sequencing runs contain pooled samples across multiple distinct experiments. To handle this, the Python script`fastq_sep_groups_v1.5.py`(an official script provided by Parse Biosciences support) was executed to parse the raw forward and reverse reads (`R1` and `R2`). This pre-filtering step segregated the high-throughput sequencing data by sorting specific sub-library wells into their target experimental groups (e.g., separating the `Astro_isol` cohort from unrelated experiment (`MCAO_BigII`)). These individual, demultiplexed FASTQ files generated by the script have been deposited and are publicly available on the NCBI database.


```{python, eval=FALSE}
python fastq_sep_groups_v1.5.py --chemistry v3 --fq1 raw/BigII_Sub01_R1_cat.fastq.gz --fq2 raw/BigII_Sub01_R2_cat.fastq.gz --output_dir split-fq/ --opref BigII_Sub01_group --group Astro_isol D4-D12 --group MCAO_BigII A1-D3

python fastq_sep_groups_v1.5.py --chemistry v3 --fq1 raw/BigII_Sub02_R1_cat.fastq.gz --fq2 raw/BigII_Sub02_R2_cat.fastq.gz --output_dir split-fq/ --opref BigII_Sub02_group --group Astro_isol D4-D12 --group MCAO_BigII A1-D3

python fastq_sep_groups_v1.5.py --chemistry v3 --fq1 raw/BigII_Sub03_R1_cat.fastq.gz --fq2 raw/BigII_Sub03_R2_cat.fastq.gz --output_dir split-fq/ --opref BigII_Sub03_group --group Astro_isol D4-D12 --group MCAO_BigII A1-D3

python fastq_sep_groups_v1.5.py --chemistry v3 --fq1 raw/BigII_Sub04_R1_cat.fastq.gz --fq2 raw/BigII_Sub04_R2_cat.fastq.gz --output_dir split-fq/ --opref BigII_Sub04_group --group Astro_isol D4-D12 --group MCAO_BigII A1-D3

python fastq_sep_groups_v1.5.py --chemistry v3 --fq1 raw/BigII_Sub05_R1_cat.fastq.gz --fq2 raw/BigII_Sub05_R2_cat.fastq.gz --output_dir split-fq/ --opref BigII_Sub05_group --group Astro_isol D4-D12 --group MCAO_BigII A1-D3

python fastq_sep_groups_v1.5.py --chemistry v3 --fq1 raw/BigII_Sub06_R1_cat.fastq.gz --fq2 raw/BigII_Sub06_R2_cat.fastq.gz --output_dir split-fq/ --opref BigII_Sub06_group --group Astro_isol D4-D12 --group MCAO_BigII A1-D3

python fastq_sep_groups_v1.5.py --chemistry v3 --fq1 raw/BigII_Sub07_R1_cat.fastq.gz --fq2 raw/BigII_Sub07_R2_cat.fastq.gz --output_dir split-fq/ --opref BigII_Sub07_group --group Astro_isol D4-D12 --group MCAO_BigII A1-D3

python fastq_sep_groups_v1.5.py --chemistry v3 --fq1 raw/BigII_Sub08_R1_cat.fastq.gz --fq2 raw/BigII_Sub08_R2_cat.fastq.gz --output_dir split-fq/ --opref BigII_Sub08_group --group Astro_isol D4-D12 --group MCAO_BigII A1-D3

python fastq_sep_groups_v1.5.py --chemistry v3 --fq1 raw/BigII_Sub09_R1_cat.fastq.gz --fq2 raw/BigII_Sub09_R2_cat.fastq.gz --output_dir split-fq/ --opref BigII_Sub09_group --group Astro_isol D4-D12 --group MCAO_BigII A1-D3

python fastq_sep_groups_v1.5.py --chemistry v3 --fq1 raw/BigII_Sub10_R1_cat.fastq.gz --fq2 raw/BigII_Sub10_R2_cat.fastq.gz --output_dir split-fq/ --opref BigII_Sub10_group --group Astro_isol D4-D12 --group MCAO_BigII A1-D3
```

################################################################
# PROCESSING THE OBTAINED FASTQ FILES VIA THE SPLIT-PIPE PIPLINE
################################################################
The following bash commands were executed in an Ubuntu environment to process the raw, pre-filtered FASTQ files using the Parse Biosciences `split-pipe` pipeline (v3 chemistry, WT kit).

```{bash, eval=FALSE}
split-pipe --mode all --chemistry v3 --kit WT --genome_dir /storage/abaffyp/Genomes/GRCm38_single_cell_ParseBio/ --fq1 ./split-fq/BigII_Sub01_group_Astro_isol_R1.fastq.gz --fq2 ./split-fq/BigII_Sub01_group_Astro_isol_R2.fastq.gz --output_dir spipe_Astro_isol/Sub01 --samp_sltab SampleLoadingTable.xlsm --nthreads 48;

split-pipe --mode all --chemistry v3 --kit WT --genome_dir /storage/abaffyp/Genomes/GRCm38_single_cell_ParseBio/ --fq1 ./split-fq/BigII_Sub02_group_Astro_isol_R1.fastq.gz --fq2 ./split-fq/BigII_Sub02_group_Astro_isol_R2.fastq.gz --output_dir spipe_Astro_isol/Sub02 --samp_sltab SampleLoadingTable.xlsm --nthreads 48;

split-pipe --mode all --chemistry v3 --kit WT --genome_dir /storage/abaffyp/Genomes/GRCm38_single_cell_ParseBio/ --fq1 ./split-fq/BigII_Sub03_group_Astro_isol_R1.fastq.gz --fq2 ./split-fq/BigII_Sub03_group_Astro_isol_R2.fastq.gz --output_dir spipe_Astro_isol/Sub03 --samp_sltab SampleLoadingTable.xlsm --nthreads 48;

split-pipe --mode all --chemistry v3 --kit WT --genome_dir /storage/abaffyp/Genomes/GRCm38_single_cell_ParseBio/ --fq1 ./split-fq/BigII_Sub04_group_Astro_isol_R1.fastq.gz --fq2 ./split-fq/BigII_Sub04_group_Astro_isol_R2.fastq.gz --output_dir spipe_Astro_isol/Sub04 --samp_sltab SampleLoadingTable.xlsm --nthreads 48;

split-pipe --mode all --chemistry v3 --kit WT --genome_dir /storage/abaffyp/Genomes/GRCm38_single_cell_ParseBio/ --fq1 ./split-fq/BigII_Sub05_group_Astro_isol_R1.fastq.gz --fq2 ./split-fq/BigII_Sub05_group_Astro_isol_R2.fastq.gz --output_dir spipe_Astro_isol/Sub05 --samp_sltab SampleLoadingTable.xlsm --nthreads 48;

split-pipe --mode all --chemistry v3 --kit WT --genome_dir /storage/abaffyp/Genomes/GRCm38_single_cell_ParseBio/ --fq1 ./split-fq/BigII_Sub06_group_Astro_isol_R1.fastq.gz --fq2 ./split-fq/BigII_Sub06_group_Astro_isol_R2.fastq.gz --output_dir spipe_Astro_isol/Sub06 --samp_sltab SampleLoadingTable.xlsm --nthreads 48;

split-pipe --mode all --chemistry v3 --kit WT --genome_dir /storage/abaffyp/Genomes/GRCm38_single_cell_ParseBio/ --fq1 ./split-fq/BigII_Sub07_group_Astro_isol_R1.fastq.gz --fq2 ./split-fq/BigII_Sub07_group_Astro_isol_R2.fastq.gz --output_dir spipe_Astro_isol/Sub07 --samp_sltab SampleLoadingTable.xlsm --nthreads 48;

split-pipe --mode all --chemistry v3 --kit WT --genome_dir /storage/abaffyp/Genomes/GRCm38_single_cell_ParseBio/ --fq1 ./split-fq/BigII_Sub08_group_Astro_isol_R1.fastq.gz --fq2 ./split-fq/BigII_Sub08_group_Astro_isol_R2.fastq.gz --output_dir spipe_Astro_isol/Sub08 --samp_sltab SampleLoadingTable.xlsm --nthreads 48;

split-pipe --mode all --chemistry v3 --kit WT --genome_dir /storage/abaffyp/Genomes/GRCm38_single_cell_ParseBio/ --fq1 ./split-fq/BigII_Sub09_group_Astro_isol_R1.fastq.gz --fq2 ./split-fq/BigII_Sub09_group_Astro_isol_R2.fastq.gz --output_dir spipe_Astro_isol/Sub09 --samp_sltab SampleLoadingTable.xlsm --nthreads 48;

split-pipe --mode all --chemistry v3 --kit WT --genome_dir /storage/abaffyp/Genomes/GRCm38_single_cell_ParseBio/ --fq1 ./split-fq/BigII_Sub10_group_Astro_isol_R1.fastq.gz --fq2 ./split-fq/BigII_Sub10_group_Astro_isol_R2.fastq.gz --output_dir spipe_Astro_isol/Sub10 --samp_sltab SampleLoadingTable.xlsm --nthreads 48;
```

##############################
# GENERATING THE SEURAT OBJECT
##############################
Following demultiplexing and initial pipeline execution via `split-pipe`, downstream data analysis was performed in an R environment. This segment of the pipeline focuses on parsing individual sample directories, executing empty droplet statistics to distinguish genuine single cells from ambient RNA background, constructing the core Seurat object, and appending comprehensive experimental metadata.

Load required package

```{r, message=FALSE, warning=FALSE, paged.print=TRUE}
suppressMessages(library("Matrix"))
suppressMessages(library("NormExpression"))
suppressMessages(library("DropletUtils"))
suppressMessages(library("DESeq2"))
suppressMessages(library("pheatmap"))
suppressMessages(library("RColorBrewer"))
suppressMessages(library("ggplot2"))
suppressMessages(library("Seurat"))
suppressMessages(library("sctransform"))
suppressMessages(library("DT"))
suppressMessages(library("scales"))
suppressMessages(library("ggrepel"))
suppressMessages(library("dplyr"))
suppressMessages(library("SeuratWrappers"))
suppressMessages(library("scmap"))
suppressMessages(library("DoubletFinder"))
suppressMessages(library("scCATCH"))
suppressMessages(library("reshape2"))
suppressMessages(library("Nebulosa"))
suppressMessages(library("clustree"))
suppressMessages(library("parallel"))
```

```{r}
sample_names <- list.dirs(path = ".", full.names = FALSE, recursive = FALSE)
```

```{r load_data, eval=FALSE, message = FALSE}
# Define a function to load data for each sample
load_sample_data <- function(sample_name) {
  folder <- paste0("./",sample_name,"/all-sample/DGE_unfiltered/")
  
  # Read matrix dimensions and the sparse matrix itself
  dims <- readr::read_delim(paste0(folder, "count_matrix.mtx"), delim = ' ', skip = 2, n_max = 1, col_names = FALSE)
  t <- readr::read_delim(paste0(folder, "count_matrix.mtx"), delim = ' ', skip = 3, col_names = FALSE)
  
  # Read barcodes and gene names
  cellbarcodes <- read.table(paste0(folder, "cell_metadata.csv"), sep = ",", header = TRUE)
  genenames <- read.table(paste0(folder, "all_genes.csv"), sep = ",", header = TRUE)
  
  # Create dimnames
  dimnames <- list(as.character(cellbarcodes$bc_wells), as.character(genenames$gene_id))
  
  # Return a list with the sparse matrix stored in the "RNA" slot
  list(RNA = t(Matrix::sparseMatrix(i = t$X1, j = t$X2, x = t$X3, dims = c(dims$X1, dims$X2), dimnames = dimnames)))
}

# Parallelize data loading using mclapply
STAR <- mclapply(sample_names, load_sample_data, mc.cores = min(nrow(samples),48))
```


```{r eval=FALSE}
# Define a function to process each sample: filter cells and generate plots
process_sample <- function(i) {
  # Initialize a list to store the QC plots
  plots <- list()
  
  # Calculate barcode ranks before filtering
  bcrank <- barcodeRanks(STAR[[i]]$RNA, lower = 700)
  uniq <- !duplicated(bcrank$rank)
  
  # Cell counts for knee and inflection points before filtering
  knee_count <- paste0("Knee = ", table(bcrank$total >= metadata(bcrank)$knee)[2], " cells")
  inflection_count <- paste0("Inflection = ", table(bcrank$total >= metadata(bcrank)$inflection)[2], " cells")
  
  # Generate QC plot before filtering
  plots[[1]] <- ggplot(data = data.frame(rank = bcrank$rank[uniq], total = bcrank$total[uniq]), aes(x = rank, y = total)) +
    geom_line() +
    geom_hline(yintercept = metadata(bcrank)$knee, color = "blue", linetype = 2) +
    geom_hline(yintercept = metadata(bcrank)$inflection, color = "darkgreen", linetype = 2) +
    annotate("text", x = max(bcrank$rank) * 0.9, y = 1.5 * c(metadata(bcrank)$knee, metadata(bcrank)$inflection),
             label = c(knee_count, inflection_count), color = c("blue", "darkgreen")) +
    scale_x_log10() + scale_y_log10() +
    labs(x = "Barcode rank", y = "Total UMI count", title = sample_names[i])
  
  # Perform emptyDrops to filter out ambient RNA
  e.out <- emptyDrops(STAR[[i]]$RNA, lower = 700, test.ambient = FALSE)
  is.cell <- e.out$FDR <= 0.01

  # Only filter and update the STAR object if cells are detected
  if (sum(is.cell, na.rm = TRUE) > 0) {
    w2kp <- which(is.cell)
    
    # Update STAR object by keeping only true cells
    STAR[[i]]$RNA <- STAR[[i]]$RNA[, w2kp]

    # Calculate barcode ranks after filtering
    bcrank <- barcodeRanks(STAR[[i]]$RNA, lower = 1000)
    uniq <- !duplicated(bcrank$rank)

    # Cell counts for knee and inflection points after filtering
    knee_count <- paste0("Knee = ", table(bcrank$total >= metadata(bcrank)$knee)[2], " cells")
    inflection_count <- paste0("Inflection = ", table(bcrank$total >= metadata(bcrank)$inflection)[2], " cells")

    # Generate QC plot after filtering
    plots[[2]] <- ggplot(data = data.frame(rank = bcrank$rank[uniq], total = bcrank$total[uniq]), aes(x = rank, y = total)) +
      geom_line() +
      geom_hline(yintercept = metadata(bcrank)$knee, color = "blue", linetype = 2) +
      geom_hline(yintercept = metadata(bcrank)$inflection, color = "darkgreen", linetype = 2) +
      annotate("text", x = max(bcrank$rank) * 0.9, y = 1.5 * c(metadata(bcrank)$knee, metadata(bcrank)$inflection),
               label = c(knee_count, inflection_count), color = c("blue", "darkgreen")) +
      scale_x_log10() + scale_y_log10() +
      labs(x = "Barcode rank", y = "Total UMI count", title = paste0(sample_names[i], " ", ncol(STAR[[i]]$RNA), " cells"))
    
    return(list(plots = plots, STAR = STAR[[i]]))  # Return both the plots and filtered STAR object
  } else {
    message("No cells detected for sample: ", sample_names[i])
    return(list(plots = plots, STAR = STAR[[i]]))  # Return original STAR object without filtering
  }
}

# Parallel processing using mclapply to process all samples in parallel
result <- parallel::mclapply(seq_along(STAR), process_sample, mc.cores = min(length(STAR),12), mc.preschedule = FALSE)  # Adjust 'mc.cores' based on available resources

# Update STAR object with filtered RNA matrices
STAR <- lapply(result, function(x) x$STAR)

# Extract and print the QC plots
for (i in seq_along(result)) {
  if (!is.null(result[[i]]$plots)) {
    for (plot in result[[i]]$plots) {
      print(plot)
    }
  }
}
```


```{r conversion, eval=FALSE, message=FALSE, warning=FALSE}
# Function to convert STAR object to Seurat and add metadata
convert_to_seurat <- function(i) {
  # Convert STAR[[i]] to Seurat object
  seurat_obj <- as.Seurat(x = STAR[[i]])
  
  # Add metadata from the samples data frame
  folder <- paste0("./",sample_names[i],"/all-sample/DGE_unfiltered/")
  cellbarcodes <- read.table(paste0(folder, "cell_metadata.csv"), sep = ",", header = TRUE)
  cellbarcodes <- cellbarcodes[cellbarcodes$bc_wells %in% colnames(seurat_obj),]
  seurat_obj@meta.data <- cbind(seurat_obj@meta.data,cellbarcodes[,-1])
  
  # Return the Seurat object
  return(seurat_obj)
}

# Parallelize the conversion of STAR objects to Seurat objects
seurat <- mclapply(seq_along(STAR), convert_to_seurat, mc.cores = min(length(STAR),12))  # Adjust mc.cores as needed

# Set names of the Seurat objects to their respective sample names
names(seurat) <- sample_names
```

```{r}
seurat <- merge(x = seurat[[1]], y = seurat[-1])

seurat <- JoinLayers(seurat)
```


```{r}
samples_2<-read.table("samples_2.csv",header=TRUE,sep=";",colClasses = c("character","character","character","character","character","character","character"))
seurat$Origin_Name <- "N"
seurat$SampleName <- "N"
seurat$well <- "N"
seurat$Condition <- "N"
seurat$Isolation <- "N"
seurat$Time <- "N"
seurat$Project <- "N"
for(j in samples_2$orig.ident){
  if(!is.na(rownames(table(rownames(table(seurat$orig.ident)) %in% j))[2])){
    seurat@meta.data[seurat$orig.ident == j,]$Origin_Name <- samples_2[samples_2$orig.ident == j, ]$Origin_Name
    seurat@meta.data[seurat$orig.ident == j,]$SampleName  <- samples_2[samples_2$orig.ident == j, ]$Sample_Name
    seurat@meta.data[seurat$orig.ident == j,]$well        <- samples_2[samples_2$orig.ident == j, ]$well
    seurat@meta.data[seurat$orig.ident == j,]$Condition   <- samples_2[samples_2$orig.ident == j, ]$Condition
    seurat@meta.data[seurat$orig.ident == j,]$Isolation   <- samples_2[samples_2$orig.ident == j, ]$Isolation
    seurat@meta.data[seurat$orig.ident == j,]$Time        <- samples_2[samples_2$orig.ident == j, ]$Time
    seurat@meta.data[seurat$orig.ident == j,]$Project     <- samples_2[samples_2$orig.ident == j, ]$Project
  }
}
```

```{r}
seurat@meta.data$Condition <- factor(seurat@meta.data$Condition, levels = c("Ctrl","MCAO","Tau"))
seurat@meta.data$Time <- factor(seurat@meta.data$Time, levels = c("Ctrl","D7","Tau"))
```


```{r}
#Create function to convert ENSMEBL_id to gene symbol and vice versa
genenames <- read.table("./Sub01/all-sample/DGE_unfiltered/all_genes.csv", header = TRUE, sep = ",") #Load file with ENSMEBL_id and gene symbol information

#Reduce genenames list only to genes detected in experiment
#genenames <- genenames[genenames$gene_id %in% rownames(seurat[[1]]),]

Symbol_to_ENSEMBL <- function(gi){
  genelist <- NULL #Create object for exporting ENSEMBL ids
  #Cycle to preserve order of genes
  for(i in 1:length(gi)){
    genelist <- c(genelist,genenames[genenames$gene_name %in% gi[i],]$gene_id) #Export ENSEMBL ids. If there is more genes with same symbol, it exports more ENSEMBL ids. If there is no, it exports nothing. You need to be carefull. If your gene symbol or ensembl id is in another column, or columns have different names, you need to change this [genenames$V2 %in% gi[i],1]
  }
  return(genelist) #Return created list
}

ENSEMBL_to_Symbol <- function(gi){
  genelist <- NULL
  for(i in 1:length(gi)){
    genelist <- c(genelist,genenames[genenames$gene_id %in% gi[i],]$gene_name)
  }
  return(genelist)
}
```


```{r calculate_mt_ribo, eval = TRUE}
#Removes ribosomal, mitochondrial genes and Malat1, which usually represent the highest proportion.
seurat_nomalat <- seurat[!(rownames(seurat) %in% Symbol_to_ENSEMBL(c(grep("^Rp[ls]",genenames$gene_name,value = TRUE),
                                                                     grep("^mt-",genenames$gene_name,value = TRUE),
                                                                     "Malat1","Gm42418","Rn18s-rs5"))), ]

seurat_nomalat$largest_count <- apply(seurat_nomalat@assays$RNA$counts, 2, max) #Find the gene whith the highest number
seurat_nomalat$largest_index <- apply(seurat_nomalat@assays$RNA$counts, 2, which.max) #Assign which gene it is
seurat_nomalat$largest_gene <- rownames(seurat_nomalat)[seurat_nomalat$largest_index] #Assing name of this gene

seurat_nomalat$percent.Largest.gene <- 100 * seurat_nomalat$largest_count / seurat_nomalat$nCount_RNA #Calculate proportion of the reads for the Largest Gene
  
#Transfer this values into original seurat object with mitochodnrial and ribosomal reads.
seurat$largest_gene <- seurat_nomalat$largest_gene
seurat$percent.Largest.gene <- seurat_nomalat$percent.Largest.gene

rm(seurat_nomalat)
gc() #clean up memory
```


Calculate the percentage of mitochondrial genes and ribosomal protein genes in each of the Seurat objects in the "seurat" list. It creates a new column in each Seurat object called "percent.mt" and "percent.rib" that contains the percentage of mitochondrial and ribosomal protein genes, respectively.
```{r calculate_mt_ribo, eval = TRUE}
seurat[["percent.mt"]] <- PercentageFeatureSet(seurat, features = Symbol_to_ENSEMBL(c(grep("^mt-",genenames$gene_name,value = TRUE))))
seurat[["percent.rib"]] <- PercentageFeatureSet(seurat, features = Symbol_to_ENSEMBL(c(grep("^Rp[ls]",genenames$gene_name,value = TRUE))))
```

##############################################
# SAVE SEURAT OBJECT FOR DOWNSTREAM PROCESSING 
##############################################
```{r eval=FALSE}
saveRDS(seurat,"seurat_Isolation_only.rds")
```